Feeding Dynamic Behavior into Optimization: Simulation-based Multi-objective Overtime Planning

### 2014-08-28 data/result/Article/Latex/Article.tex
Atualizando o latex

\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}
\hyphenation{op-tical net-works semi-conduc-tor}

@@

\author{\IEEEauthorblockN{Márcio de O. Barros\IEEEauthorrefmark{1},
						  Luiz Antonio O. de Araujo Jr\IEEEauthorrefmark{2}
        \IEEEauthorblockA{Federal University of the State of Rio de Janeiro\\
Av. Pasteur 458, CEP 22.290-240
Rio de Janeiro, Brazil
\\ \{marcio.barros\IEEEauthorrefmark{1}, luiz.araujojr\IEEEauthorrefmark{2}\}@uniriotec.br}
}}

@@

Software engineers are professionals particularly subjected to working overtime, frequently without previous planning and in stressful, fire-fighting situations \cite{Frangos:1997}, \cite{Hyman:2003}, \cite{Houdmont:2011}. Difficulties associated to measuring the progress of a software project, late and volatile requirements, compressed schedules, and the 
 to shrink the time-to-market of new features \cite{Halliday:2011} are among the factors that contribute to the need of working overtime.

@@

Problems related to overtime are not exclusive to the IT sector. Caruso \cite{caruso:2006} identified negative impacts of long work hours on American workers (fatigue, lack of alertness, weight gain, and health degradation), their families (delayed marriage and childbearing), and their employers (absenteeism due to illness, increased errors, and rework). The author also reminds about the extra cost incurred in order to identify and correct errors committed due to exhaustion (loss in quantity and quality of the goods produced and services provided by workers subjected to excessive overtime). Donald et al. \cite{Donald:2005} also found negative correlation between stress and productivity. Finally, overtime problems do not seem related to a specific culture, affecting Americans \cite{caruso:2006}, \cite{Dembe:2005}, Brazilians \cite{fischer:2000}, Japanese \cite{karita:2006}, \cite{Sasaki:2007}, and Chinese \cite{Houdmont:2011}, to cite a few.

@@

We depart from a set of work packages, each to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP = \{wp1, wp2, ..., wpn\} of work packages and an edge set WPDEP = \{(wpi, wpj): i$\not=$j,1$\leq$i$\leq$n,1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each wpi is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package (wpi) whose development can only start after the requirements for a second work package (wpj) are known.

From these work packages, we develop a project schedule. Our schedule is a partial order of software development activities on which four types of activities are created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes after a requirement analysis activity. These dependencies represent the basic workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package wi cannot start before the requirements analysis activity for wj is concluded for all (wi, wj) pairs in WPDEP.

A project schedule is represented as an acyclic directed graph $<$ACT, ACDEP$>$ consisting of a node set ACT = \{a1, a2, ..., an\} of activities and an edge set ACDEP = \{(ai, aj): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity ai is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce its results and the expected number of errors they might generate if performed by a mean developer in a regular environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information provided in work packages.
We consider three objectives for overtime planning: the number of overtime hours (NOH), project makespan (MKS), and overall project cost (CST). OPP is formulated as a three objective decision problem in which makespan, cost, and overtime hours are conflicting objectives to be minimized. A candidate solution for the OPP is a sequence of numbers that represents the amount of daily overtime hours to be spent on each activity comprising the schedule. Our formulation is based on a 40-hours work week, accepts multiples of 30 minutes for overtime work, and limits the maximum amount of daily work to 12-hours turns, according to Brazilian law. Thus, for any given activity we may assign zero overtime, 30 minutes of overtime, one hour, one hour and 30 minutes, and so on up to four hours of daily overtime. Our first objective, NOH, is calculated as the sum of the amount of overtime hours assigned for each activity in the schedule.

@@

The last objective, project cost (CST), is the sum of each activity’s cost. The cost of an activity depends on the number of regular work hours and overtime hours its developers consume to produce its expected results and is related to the duration calculated by the simulator. Again, we modeled the problem according to Brazilian laws: if a regular working hour costs X, each of the first two overtime hours costs 120\%.X and the next two overtime hours cost 150\%.X each.

\subsection{Simulation Machanism}
Simulation is the imitation of the behavior presented by a real-world process or system over time \cite{Banks:2000}. It supports predicting and understanding the behavior of complex systems, containing hundreds or thousands of interconnected components. Simulation is useful because the number of components and the complexity of the data and control exchange among them usually hampers our ability to interpret system behavior by intuition or analytical reasoning.

@@

Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. The internal state of an object is kept in its instance variables. Internal state adjustment operations are coded in methods activated by the simulator: init(), to initialize the simulation object at time zero; start(), called when all dependencies for an object are finished; step(), called at each simulation step; and finish(), called when the object signals its conclusion.

@@

Scenarios are used to represent complementary behavior that may be imposed to certain simulation objects. A scenario is written to influence a certain type of simulation object and only participates in the simulation if activated on one or more objects of that type. Scenarios can query and update the internal state of their associated objects before and after each event on which the objects update their internal state (start, step, finish). Scenarios can be used to represent managerial decisions imposed to a subset of a software project's activities. They are reusable model components that modify the behavior of their related simulation object according to the policy they represent. For the purpose of the OPP, a scenario was developed to represent the effects of overtime while a base model describes the expected behavior for a project, its activities and developers as if uninfluenced by managerial policies and decisions (such as overtime allocation).

@@

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to de Marco’s law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50% more errors than while working on a regular shift.

@@

\section{Analysis and Discussion}
Below we present results obtained from our experiments to support answering the proposed research questions. \linebreak
\textbf{RQ1:} To compare NSGAII to random search (RS), we ran 50 independent optimization cycles for both algorithms using the same stop criteria. NSGAII achieved higher I$_{HV}$, higher I$_{C}$, and lower  on all six instances compared to RS. Wilcoxon-Mann-Whitney tests applied on an instance basis found significant differences between results produced by the algorithms at 95\% significance level for all quality indicators and all instances. Standardized effect-sizes (non-parametric Vargha-Delaney Â12 test) also indicate that NSGAII produces higher I$_{HV}$ and lower  in 100\% of the runs, while producing higher I$_{C}$ for more than 88\% of these runs. Thus, there is strong evidence that NSGAII outperforms RS while searching for solutions for the OPP and that our formulation and parameterization passes the "sanity check" implied in the first research question.\linebreak
\textbf{RQ2:} To address this question we compared our formulation for the OPP with three overtime management strategies (OMS) presented by Ferrucci et al. \cite{Ferrucci:2013} as reference practices from the industry: (i) "margarine" management (MAR) suggests spreading overtime hours evenly over all activities comprising the project schedule; (ii) Critical Path (CPM) suggests loading overtime onto the schedule's critical path to reduce completion time and risk of schedule overrun; and (iii) Second Half (SH) suggests loading overtime hours onto the later part of the project's schedule to compensate for delays introduced in early activities.

@@

The Â12 standardized effect-size was calculated for pair-wise comparisons between NSGAII and each OMS for all quality indicators. Results are presented in Table III, denoting the chances of NSGAII to produce higher IC and I$_{HV}$ and lower  than the related OMS. Â12 is very high and favors NSGAII in almost all instances for I$_{HV}$ and  (except PARM for I$_{HV}$, OMET and PARM for I$_{GD}$). It strongly favors NSGAII in terms of IC if compared to MAR and CPM, except on instance PARM (a draw on effect-size among both OMS and NSGAII). On the other hand, Â12 tells a different story for I$_{C}$ between NSGAII and SH, clearly favoring SH for almost all instances.

@@

% Table generated by Excel2LaTeX from sheet 'tabela 4'
\begin{table}[htbp]
  \centering
  \caption{Quality indicators for each instance under our approach (NSGAII) and a model without error dynamics (NSGA$_{NE}$). NSGA$_{NE}$ produces Pareto fronts that cover a larger part of the search space, leading to smalled . Solutions produced under NSGA-II tend to concentrate in certain parts of the search space}
    \begin{tabular}{p{0.8cm}p{1.0cm}ccc}
    \toprule
    \textbf{Instance} & \textbf{Config.} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{I$_{GD}$} \\
    \midrule
    \multirow{2}[2]{*}{ACAD} & NSGAII & 0.020 ± 0.022 & \textbf{0.464 ± 0.009} & 0.0066 ± 0.002 \\
          & NSGA$_{NE}$ & \textbf{0.020 ± 0.004} & 0.333 ± 0.001 & \textbf{0.0003 ± 0.0} \\
    \multirow{2}[2]{*}{WMET} & NSGAII & 0.020 ± 0.019 & \textbf{0.492 ± 0.006} & 0.0037 ± 0.0 \\
          & NSGA$_{NE}$ & 0.020 ± 0.003 & 0.334 ± 0.001 & \textbf{0.0003 ± 0.0} \\
    \multirow{2}[2]{*}{WAMS} & NSGAII & 0.0200 ± 0.016 & \textbf{0.479 ± 0.004} & 0.0032 ± 0.0 \\
          & NSGA$_{NE}$ & 0.0200 ± 0.003 & 0.337 ± 0.001 & \textbf{0.0002 ± 0.0} \\
    \multirow{2}[2]{*}{PSOA} & NSGAII & 0.020 ± 0.014 & \textbf{0.421 ± 0.014} & 0.0129 ± 0.004 \\
          & NSGA$_{NE}$ & 0.020 ± 0.002 & 0.338 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \multirow{2}[2]{*}{OMET} & NSGAII & 0.020 ± 0.014 & \textbf{0.465 ± 0.006} & 0.0044 ± 0.0 \\
          & NSGA$_{NE}$ & 0.020 ± 0.003 & 0.336 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \multirow{2}[2]{*}{PARM} & NSGAII & 0.020 ± 0.019 & \textbf{0.489 ± 0.011} & 0.0090 ± 0.001 \\
          & NSGA$_{NE}$ & \textbf{0.020 ± 0.003} & 0.338 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \bottomrule
    \end{tabular}%
\end{table}%

% Table generated by Excel2LaTeX from sheet 'tabela 5'
\begin{table}[htbp]
  \centering
  \caption{Effect-sizes for comparisons between NSGAII and NSGA$_{NE}$}
    \begin{tabular}{lccc}
    \toprule
    \textbf{Instance} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{} \\
    \midrule
    ACAD  & 38\%  & 100\% & 0\% \\
    WMET  & 40\%  & 100\% & 0\% \\
    WAMS  & 46\%  & 100\% & 0\% \\
    PSOA  & 50\%  & 100\% & 0\% \\
    OMET  & 43\%  & 100\% & 0\% \\
    PARM  & 38\%  & 100\% & 0\% \\
    \bottomrule
    \end{tabular}%
\end{table}%

Â12 effect-sizes for each quality indicator are presented in Table V. It can be observed that NSGAII consistently presents higher I$_{HV}$ (good) and higher  (bad) compared to NSGA$_{NE}$. On regard of I$_{C}$, the algorithms are competitive, with small effect-sizes favoring NSGA$_{NE}$. These results are explained by the larger number of solutions produced by NSGA$_{NE}$: its Pareto fronts have on average 1.8 times the number of solutions of fronts calculated for NSGAII, reaching up to 3.9 times this number for the largest instance (PARM). The larger number of solutions allows NSGA$_{NE}$ to cover parts of the reference front uncovered by NSGAII, lowering its  and increasing its I$_{C}$. NSGAII contributes with more points than NSGA$_{NE}$ on parts of the reference front that it covers, but these improvements are not sufficient to compensate for the distance between points in the extremes of its fronts and the parts of the reference front defined only by NSGA$_{NE}$.

% Table generated by Excel2LaTeX from sheet 'tabela 6'
\begin{table}[htbp]
  \centering
  \caption{Average difference in makespan and cost for solutions involving excessive overtime for each instance. data shows that the formulation without overtime error dynamics tends to understimate both cost and makespan for most projects.}
    \begin{tabular}{rccc}
    \toprule
    \multicolumn{1}{c}{\multirow{2}[2]{*}{\textbf{Instance}}} & \multirow{2}[2]{*}{\textbf{Overtime  Range}} & \multicolumn{2}{c}{\textbf{(NSGA$_{NE}>$NSGA-II)}} \\

    \multicolumn{1}{c}{} &       & \textbf{Makespan } & \textbf{Cost} \\
        \midrule
    \multicolumn{1}{l}{ACAD} & 332 - 376 hr & -2\%  & -1\% \\
    \multicolumn{1}{l}{WMET} & 332 - 372 hr & -4\%  & -3\% \\
    \multicolumn{1}{l}{WAMS} & 570 - 638 hr & -5\%  & -3\% \\
    \multicolumn{1}{l}{PSOA} & 1,288 - 1,443 hr & 6\%   & 5\% \\
    \multicolumn{1}{l}{OMET} & 919 - 1,023 hr & -6\%  & -4\% \\
    \multicolumn{1}{l}{PARM} & 797 - 884 hr & -5\%  & -3\% \\
    \bottomrule
    \end{tabular}%
\end{table}%
Besides quality indicators, it is interesting to consider the differences observed in objective values produced under the formulations with and without overtime error dynamics. Table VI presents the differences in average makespan and cost observed in solutions involving the highest overtime allocation for each instance (about 3 hours of daily overtime over the project life-cycle). Data shows that NSGA$_{NE}$ tends to underestimate project cost and makespan when a large number of overtime hours are imposed to the development team. Such difference, observed in 5 out of 6 instances, is due to longer testing activities required to identify and correct the exceeding errors introduced in the software by tired developers.
Based on the analysis above, we have found evidence that optimizing overtime planning without taking into account an increase in error generation rates due to developers working overtime produces distinct results from an optimization process that considers such an increase. Next, we explore the practical implications of these findings.
Lessons Learned: Insights for the decision maker are better appreciated in graphical format. Figure II presents 2D projections of pair-wise combinations for the three objectives comprising our formulation for the OPP. Each chart presents the best Pareto front over 50 optimization cycles under NSGAII (circles), the best Pareto front over 50 cycles under NSGA$_{NE}$ (line), and Pareto fronts for CPM (), MAR (x), and SH (). Results are shown for the smallest and the largest instances in our sample (respectively, ACAD and PARM).
%simbolos
A first result that can be observed from these charts is the linear shape of the Pareto front calculated for NSGA$_{NE}$. This front is comprised of a set of points distributed along the main diagonal of the hypercube representing the objective space for the OPP. The shape of the Pareto front implies that there exist linear relationships between the number of overtime hours invested on the project, project makespan and cost if the dynamics in error generation rates is not taken into account. These linear relationships are incompatible with previous findings in Software Engineering research, which postulate an exponential relationship between cost (effort) and duration for non-trivial software projects \cite{Boehm:1981}. They may also make the optimization process easier for the genetic algorithm, resulting in a larger pool of non-dominated solutions and probing parts of the objective space which could not be explored by the more complex search accounting for error dynamics.

\begin{figure*}
\centering
\includegraphics[width=18cm]{lessonslearned.jpg}
\caption{2d projections for combinations of the number of overtime hours, makespan, and cost objectives for the smaller (acad) and largest (parm) instances. the linear relations resulting from NSGA$_{NE}$ and the convex shape of nsga-ii can be observed in the charts, as well as the few solutions provided by each oms.}
\end{figure*}

@@

Knee-points connecting different regions of the Pareto front (such as the one described above) can be explored by decision-makers' as overtime allocation alternatives on which large changes in one objective can be garnered by accepting small compromises on other objectives. This allows decision-makers to answer questions such as "what must I do to reduce makespan?", "how much it costs to reduce makespan by X days?" or "which activities provide the best leverage for a certain amount of overtime?".

>>>>

\usepackage{bigstrut}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{booktabs}

\hyphenation{op-tical net-works semi-conduc-tor}

@@

\author{\IEEEauthorblockN{M\'arcio de O. Barros\IEEEauthorrefmark{1},
						  Luiz Antonio O. de Araujo Jr\IEEEauthorrefmark{2}
        \IEEEauthorblockA{Federal University of the State of Rio de Janeiro\\
Av. Pasteur 458, CEP 22.290-240
Rio de Janeiro, Brazil
\\ \{marcio.barros\IEEEauthorrefmark{1}, luiz.araujojr\IEEEauthorrefmark{2}\}@uniriotec.br}
}}

@@

Software engineers are professionals particularly subjected to working overtime, frequently without previous planning and in stressful, fire-fighting situations \cite{Frangos:1997}\cite{Hyman:2003}\cite{Houdmont:2011}. Difficulties associated to measuring the progress of a software project, late and volatile requirements, compressed schedules, and the 
 to shrink the time-to-market of new features \cite{Halliday:2011} are among the factors that contribute to the need of working overtime.

@@

Problems related to overtime are not exclusive to the IT sector. Caruso \cite{caruso:2006} identified negative impacts of long work hours on American workers (fatigue, lack of alertness, weight gain, and health degradation), their families (delayed marriage and childbearing), and their employers (absenteeism due to illness, increased errors, and rework). The author also reminds about the extra cost incurred in order to identify and correct errors committed due to exhaustion (loss in quantity and quality of the goods produced and services provided by workers subjected to excessive overtime). Donald et al. \cite{Donald:2005} also found negative correlation between stress and productivity. Finally, overtime problems do not seem related to a specific culture, affecting Americans \cite{caruso:2006}\cite{Dembe:2005}, Brazilians \cite{fischer:2000}, Japanese \cite{karita:2006}, \cite{Sasaki:2007}, and Chinese \cite{Houdmont:2011}, to cite a few.

@@

We depart from a set of work packages, each to be analyzed, designed, coded, and tested. A project is represented as an acyclic directed graph $<$WP, WPDEP$>$ consisting of a node set WP = \{wp$_{1}$, wp$_{2}$, ..., wp$_{n}$\} of work packages and an edge set WPDEP = \{(wp$_{i}$, wp$_{j}$): i$\not=$j,1$\leq$i$\leq$n,1$\leq$j$\leq$n\} of precedence dependencies amongst work packages. Each wp$_{i}$ is described by the amount of effort required to elicit its requirements, the amount of effort to design it, code it, and test it as if these tasks were performed by a mean developer in a regular work environment. A work package is also described by the number of errors its requirements analysis, design, and coding activities might generate in a similar context. On Section II.B we discuss how these parameters were filled for our instances. Each dependence in WPDEP represents a work package (wp$_{i}$) whose development can only start after the requirements for a second work package (wp$_{j}$) are known.

From these work packages, we develop a project schedule. Our schedule is a partial order of software development activities on which four types of activities are created for each work package: requirement analysis, design, coding, and testing. A testing activity is always preceded by a coding activity, which is preceded by a design activity, which comes after a requirement analysis activity. These dependencies represent the basic workflow of modeling the problem, designing the solution, implementing its source-code, and testing its features. The schedule also includes dependencies amongst work packages: the requirements analysis activity for work package wp$_{i}$ cannot start before the requirements analysis activity for wp$_{j}$ is concluded for all (wp$_{i}$, wp$_{j}$) pairs in WPDEP.

A project schedule is represented as an acyclic directed graph $<$ACT, ACDEP$>$ consisting of a node set ACT = \{a$_{1}$, a$_{2}$, ..., an\} of activities and an edge set ACDEP = \{(a$_{i}$, a$_{j}$): i$\not=$j,1$\leq$i$\leq$m,1$\leq$j$\leq$m\} of finish-start (FS) dependencies amongst activities. A FS dependency A$\rightarrow$B indicates that work on activity B can only start when activity A is finished. Each activity a$_{i}$ is classified as either requirement analysis, design, coding, or testing. Analysis, design, and coding activities are described by the expected effort required to produce its results and the expected number of errors they might generate if performed by a mean developer in a regular environment. Testing activities are described by the average effort required to identify and correct an error, which is collected from information provided in work packages.
We consider three objectives for overtime planning: the number of overtime hours (NOH), project makespan (MKS), and overall project cost (CST). OPP is formulated as a three objective decision problem in which makespan, cost, and overtime hours are conflicting objectives to be minimized. A candidate solution for the OPP is a sequence of numbers that represents the amount of daily overtime hours to be spent on each activity comprising the schedule. Our formulation is based on a 40-hours work week, accepts multiples of 30 minutes for overtime work, and limits the maximum amount of daily work to 12-hours turns, according to Brazilian law. Thus, for any given activity we may assign zero overtime, 30 minutes of overtime, one hour, one hour and 30 minutes, and so on up to four hours of daily overtime. Our first objective, NOH, is calculated as the sum of the amount of overtime hours assigned for each activity in the schedule.

@@

The last objective, project cost (CST), is the sum of each activity's cost. The cost of an activity depends on the number of regular work hours and overtime hours its developers consume to produce its expected results and is related to the duration calculated by the simulator. Again, we modeled the problem according to Brazilian laws: if a regular working hour costs X, each of the first two overtime hours costs 120\%.X and the next two overtime hours cost 150\%.X each.

\subsection{Simulation Machanism}
Simulation is the imitation of the behavior presented by a real-world process or system over time \cite{Banks:2000}. It supports predicting and understanding the behavior of complex systems, containing hundreds or thousands of interconnected components. Simulation is useful because the number of components and the complexity of the data and control exchange among them usually hampers our ability to interpret system behavior by intuition or analytical reasoning.

@@

Simulation objects are Java objects that implement an interface defining a protocol through which they are controlled by the simulator. The internal state of an object is kept in its instance variables. Internal state adjustment operations are coded in methods activated by the simulator: \textit{init()}, to initialize the simulation object at time zero; \textit{start()}, called when all dependencies for an object are finished; \textit{step()}, called at each simulation step; and \textit{finish()}, called when the object signals its conclusion.

@@

\textit{Scenarios} are used to represent complementary behavior that may be imposed to certain simulation objects. A scenario is written to influence a certain type of simulation object and only participates in the simulation if activated on one or more objects of that type. Scenarios can query and update the internal state of their associated objects before and after each event on which the objects update their internal state (\textit{start}, \textit{step}, \textit{finish}). Scenarios can be used to represent managerial decisions imposed to a subset of a software project's activities. They are reusable model components that modify the behavior of their related simulation object according to the policy they represent. For the purpose of the OPP, a scenario was developed to represent the effects of overtime while a base model describes the expected behavior for a project, its activities and developers as if uninfluenced by managerial policies and decisions (such as overtime allocation).

@@

The scenario used in our simulation describes the dynamics of working overtime and its influence on developer's productivity and error generation rates. If a developer is expected to work more than 8 hours per day in a given activity, his productivity will be increased proportionally to the extra number of work hours, reaching up to a 50\% increase if working to the upper limit of 12 hours per day. However, according to de Marco’s law \cite{demarco:1999}, a developer working more than the regular daily hours may produce software below the quality standard. The more hours a developer works, the more tired he becomes. Tired developers become less careful to the work they are performing and, consequently, they tend to introduce more errors in the software. We have used data from \cite{Abdel-Hamid:1991}  to estimate the increase in the number of errors produced according to the number of hours worked by a developer. Such data suggest that software quality tends to deteriorate if developers work more than 9 hours/day and that a developer working 12 hours/day commits 50\% more errors than while working on a regular shift.

@@

\section{Analysis and Discussion}
Below we present results obtained from our experiments to support answering the proposed research questions. \linebreak
\textbf{RQ1:} To compare NSGAII to random search (RS), we ran 50 independent optimization cycles for both algorithms using the same stop criteria. NSGAII achieved higher I$_{HV}$, higher I$_{C}$, and lower  on all six instances compared to RS. Wilcoxon-Mann-Whitney tests applied on an instance basis found significant differences between results produced by the algorithms at 95\% significance level for all quality indicators and all instances. Standardized effect-sizes (non-parametric Vargha-Delaney \^A12 test) also indicate that NSGAII produces higher I$_{HV}$ and lower  in 100\% of the runs, while producing higher I$_{C}$ for more than 88\% of these runs. Thus, there is strong evidence that NSGAII outperforms RS while searching for solutions for the OPP and that our formulation and parameterization passes the "sanity check" implied in the first research question.\linebreak
\textbf{RQ2:} To address this question we compared our formulation for the OPP with three overtime management strategies (OMS) presented by Ferrucci et al. \cite{Ferrucci:2013} as reference practices from the industry: (i) "margarine" management (MAR) suggests spreading overtime hours evenly over all activities comprising the project schedule; (ii) Critical Path (CPM) suggests loading overtime onto the schedule's critical path to reduce completion time and risk of schedule overrun; and (iii) Second Half (SH) suggests loading overtime hours onto the later part of the project's schedule to compensate for delays introduced in early activities.

@@

The \^A12 standardized effect-size was calculated for pair-wise comparisons between NSGAII and each OMS for all quality indicators. Results are presented in Table III, denoting the chances of NSGAII to produce higher IC and I$_{HV}$ and lower  than the related OMS. \^A12 is very high and favors NSGAII in almost all instances for I$_{HV}$ and  (except PARM for I$_{HV}$, OMET and PARM for I$_{GD}$). It strongly favors NSGAII in terms of IC if compared to MAR and CPM, except on instance PARM (a draw on effect-size among both OMS and NSGAII). On the other hand, \^A12 tells a different story for I$_{C}$ between NSGAII and SH, clearly favoring SH for almost all instances.

@@

% Table generated by Excel2LaTeX from sheet 'tabela 4'
\begin{table}[htbp]
  \centering
  \caption{Quality indicators for each instance under our approach (NSGAII) and a model without error dynamics (NSGA$_{NE}$). NSGA$_{NE}$ produces Pareto fronts that cover a larger part of the search space, leading to smalled . Solutions produced under NSGA-II tend to concentrate in certain parts of the search space}
    \begin{tabular}{p{0.8cm}p{1.0cm}ccc}
    \toprule
    \textbf{Instance} & \textbf{Config.} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{I$_{GD}$} \\
    \midrule
    \multirow{2}[2]{*}{ACAD} & NSGAII & 0.020 ± 0.022 & \textbf{0.464 ± 0.009} & 0.0066 ± 0.002 \\
          & NSGA$_{NE}$ & \textbf{0.020 ± 0.004} & 0.333 ± 0.001 & \textbf{0.0003 ± 0.0} \\
    \multirow{2}[2]{*}{WMET} & NSGAII & 0.020 ± 0.019 & \textbf{0.492 ± 0.006} & 0.0037 ± 0.0 \\
          & NSGA$_{NE}$ & 0.020 ± 0.003 & 0.334 ± 0.001 & \textbf{0.0003 ± 0.0} \\
    \multirow{2}[2]{*}{WAMS} & NSGAII & 0.0200 ± 0.016 & \textbf{0.479 ± 0.004} & 0.0032 ± 0.0 \\
          & NSGA$_{NE}$ & 0.0200 ± 0.003 & 0.337 ± 0.001 & \textbf{0.0002 ± 0.0} \\
    \multirow{2}[2]{*}{PSOA} & NSGAII & 0.020 ± 0.014 & \textbf{0.421 ± 0.014} & 0.0129 ± 0.004 \\
          & NSGA$_{NE}$ & 0.020 ± 0.002 & 0.338 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \multirow{2}[2]{*}{OMET} & NSGAII & 0.020 ± 0.014 & \textbf{0.465 ± 0.006} & 0.0044 ± 0.0 \\
          & NSGA$_{NE}$ & 0.020 ± 0.003 & 0.336 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \multirow{2}[2]{*}{PARM} & NSGAII & 0.020 ± 0.019 & \textbf{0.489 ± 0.011} & 0.0090 ± 0.001 \\
          & NSGA$_{NE}$ & \textbf{0.020 ± 0.003} & 0.338 ± 0.001 & \textbf{0.0001 ± 0.0} \\
    \bottomrule
    \end{tabular}%
\end{table}%
% Table generated by Excel2LaTeX from sheet 'tabela 5'
\begin{table}[htbp]
  \centering
  \caption{Effect-sizes for comparisons between NSGAII and NSGA$_{NE}$}
    \begin{tabular}{lccc}
    \toprule
    \textbf{Instance} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{} \\
    \midrule
    ACAD  & 38\%  & 100\% & 0\% \\
    WMET  & 40\%  & 100\% & 0\% \\
    WAMS  & 46\%  & 100\% & 0\% \\
    PSOA  & 50\%  & 100\% & 0\% \\
    OMET  & 43\%  & 100\% & 0\% \\
    PARM  & 38\%  & 100\% & 0\% \\
    \bottomrule
    \end{tabular}%
\end{table}%
\^A12 effect-sizes for each quality indicator are presented in Table V. It can be observed that NSGAII consistently presents higher I$_{HV}$ (good) and higher  (bad) compared to NSGA$_{NE}$. On regard of I$_{C}$, the algorithms are competitive, with small effect-sizes favoring NSGA$_{NE}$. These results are explained by the larger number of solutions produced by NSGA$_{NE}$: its Pareto fronts have on average 1.8 times the number of solutions of fronts calculated for NSGAII, reaching up to 3.9 times this number for the largest instance (PARM). The larger number of solutions allows NSGA$_{NE}$ to cover parts of the reference front uncovered by NSGAII, lowering its  and increasing its I$_{C}$. NSGAII contributes with more points than NSGA$_{NE}$ on parts of the reference front that it covers, but these improvements are not sufficient to compensate for the distance between points in the extremes of its fronts and the parts of the reference front defined only by NSGA$_{NE}$.
% Table generated by Excel2LaTeX from sheet 'tabela 6'
\begin{table}[htbp]
  \centering
  \caption{Average difference in makespan and cost for solutions involving excessive overtime for each instance. data shows that the formulation without overtime error dynamics tends to understimate both cost and makespan for most projects.}
    \begin{tabular}{rccc}
    \toprule
    \multicolumn{1}{c}{\multirow{2}[2]{*}{\textbf{Instance}}} & \multirow{2}[2]{*}{\textbf{Overtime  Range}} & \multicolumn{2}{c}{\textbf{(NSGA$_{NE}>$NSGA-II)}} \\

    \multicolumn{1}{c}{} &       & \textbf{Makespan } & \textbf{Cost} \\
        \midrule
    \multicolumn{1}{l}{ACAD} & 332 - 376 hr & -2\%  & -1\% \\
    \multicolumn{1}{l}{WMET} & 332 - 372 hr & -4\%  & -3\% \\
    \multicolumn{1}{l}{WAMS} & 570 - 638 hr & -5\%  & -3\% \\
    \multicolumn{1}{l}{PSOA} & 1,288 - 1,443 hr & 6\%   & 5\% \\
    \multicolumn{1}{l}{OMET} & 919 - 1,023 hr & -6\%  & -4\% \\
    \multicolumn{1}{l}{PARM} & 797 - 884 hr & -5\%  & -3\% \\
    \bottomrule
    \end{tabular}%
\end{table}%
Besides quality indicators, it is interesting to consider the differences observed in objective values produced under the formulations with and without overtime error dynamics. Table VI presents the differences in average makespan and cost observed in solutions involving the highest overtime allocation for each instance (about 3 hours of daily overtime over the project life-cycle). Data shows that NSGA$_{NE}$ tends to underestimate project cost and makespan when a large number of overtime hours are imposed to the development team. Such difference, observed in 5 out of 6 instances, is due to longer testing activities required to identify and correct the exceeding errors introduced in the software by tired developers.
Based on the analysis above, we have found evidence that optimizing overtime planning without taking into account an increase in error generation rates due to developers working overtime produces distinct results from an optimization process that considers such an increase. Next, we explore the practical implications of these findings.

\textbf{Lessons Learned:} Insights for the decision maker are better appreciated in graphical format. Figure II presents 2D projections of pair-wise combinations for the three objectives comprising our formulation for the OPP. Each chart presents the best Pareto front over 50 optimization cycles under NSGAII (circles), the best Pareto front over 50 cycles under NSGA$_{NE}$ (line), and Pareto fronts for CPM (+), MAR (x), and SH (*). Results are shown for the smallest and the largest instances in our sample (respectively, ACAD and PARM).
%simbolos
A first result that can be observed from these charts is the linear shape of the Pareto front calculated for NSGA$_{NE}$. This front is comprised of a set of points distributed along the main diagonal of the hypercube representing the objective space for the OPP. The shape of the Pareto front implies that there exist linear relationships between the number of overtime hours invested on the project, project makespan and cost if the dynamics in error generation rates is not taken into account. These linear relationships are incompatible with previous findings in Software Engineering research, which postulate an exponential relationship between cost (effort) and duration for non-trivial software projects \cite{Boehm:1981}. They may also make the optimization process easier for the genetic algorithm, resulting in a larger pool of non-dominated solutions and probing parts of the objective space which could not be explored by the more complex search accounting for error dynamics.

\begin{figure*}
\centering
\includegraphics[width=18cm]{lessonslearned.jpg}
\caption{2D projections for combinations of the number of overtime hours, makespan, and cost objectives for the smaller (acad) and largest (parm) instances. the linear relations resulting from NSGA$_{NE}$ and the convex shape of nsga-ii can be observed in the charts, as well as the few solutions provided by each oms.}
\end{figure*}

@@

Knee-points connecting different regions of the Pareto front (such as the one described above) can be explored by decision-makers' as overtime allocation alternatives on which large changes in one objective can be garnered by accepting small compromises on other objectives. This allows decision-makers to answer questions such as \textit{"what must I do to reduce makespan?", "how much it costs to reduce makespan by X days?" or "which activities provide the best leverage for a certain amount of overtime?"}.
